---
layout: manualsection
title: Loading HybridCheck and Entering Sequence Data
permalink: 02-manual.html
manual: true
published: false
status: process
---

Starting HybridCheck and Loading Data
------------------------------------

To get started fire up the R condole and enter:

```{r}
library(HybridCheck)
```

You should see the above result in the console:

The package is now successfully loaded.

---

Loading in sequence data and starting a new HybridCheck analysis
------------------------------------------------------------

To get started using HybridCheck to analyse a set of sequences, you first have to create a new HybridCheck object.
 
The HybridCheck object will contain all data, options, and will carry out the analysis. So everything you do in HybridCheck is done by interacting with this object.
By providing a file-path to a FASTA format sequence file when the new object is created the DNA data will also be loaded and prepared automatically.

```{r}
MyAnalysis <- HC$new("~/Dropbox/MySequence.fas")
```
The new HybridCheck object is ssigned to a variable "MyAnalysis".

HybridCheck prints several progress messages and let's you know when the object has been successfully created:

```R
Detected file is supposed to be a FASTA format file...
Reading in DNA sequence file...
Looking for duplicates...
Done...
Subsetting the informative segregating sites...
Done, now saving data internally
 :Full Sequence
 :Informative bases only
Finished DNA input.
```

The HybridCheck package has been loaded, and data has been read in and is now contained in a HybridCheck object, which also contains all analysis settings and steps.