---
layout: manualsection
title: Loading HybRIDS and Entering Sequence Data
permalink: 02-manual.html
manual: true
published: false
status: process
---

Starting HybRIDS and Loading Data
------------------------------------

To get started fire up the R condole and enter:

```{r}
library(HybRIDS)
```

You should see the above result in the console:

The package is now successfully loaded.

---

Loading in sequence data and starting a new HybRIDS analysis
------------------------------------------------------------

To get started using HybRIDS to analyse a set of sequences, you first have to create a new HybRIDS object. Do this either with the R console or the Graphical Interface.

**With the R console:**
 
The HybRIDS object will contain all data, options, and will carry out the analysis. So everything you do in HybRIDS is done by interacting with this object.
By providing a file-path to a FASTA format sequence file when the new object is created the DNA data will also be loaded and prepared automatically.

```{r}
MyAnalysis <- HybRIDS$new("~/Desktop/MySequence.fasta")
```
With the above line I have created a new HybRIDS object and assigned it to a variable "MyAnalysis" - that is to say I made a new HybRIDS object that is called "MyAnalysis" and now it is in the R workspace ready for us to use by calling it's methods.

The HybRIDS package has been loaded, and data has been read in and is now contained in a HybRIDS object, which also contains all analysis settings and steps.