---
layout: manualsection
title: Scanning triplets for recombination signal
permalink: 06-manual.html
manual: true
published: false
status: process
---

```{r include=FALSE, echo=FALSE}
FirstGroup <- c("Seq1", "Seq2", "Seq3")
SecondGroup <- c("Seq4", "Seq5", "Seq6")
ThirdGroup <- c("Seq7", "Seq8", "Seq9", "Seq10")
myGroups <- list(FirstGroup, SecondGroup, ThirdGroup) # Put the groups in a list.
MyAnalysis$setParameters("TripletGeneration", Groups = myGroups, PartitionStrictness = 1L)
MyAnalysis
```

This step can be started by using the `analyzeSS()` method of the HybRIDS object.
As a reminder the parameters of the step are:

WindowSize
  : The size in bases of the Sliding Window. The default size is 100.

StepSize
  : The number of bases that the sliding window moves across sequences. The default size is 1. 

The `analyzeSS` method also accepts an argument which tells it which triplets
(of the set you have defined in the `TripletGeneration` step) are to be scanned with the current
`SSAnalysis` settings.

This argument can be set as `"NOT.SCANNED"` (default) in which case, every triplet of the set defined by the 
`TripletGeneration` parameters will be scanned for recombination signal with the current `SSAnalysis`
settings, if it has not previously already been scanned.
This argument can also be set to `"ALL"`, in which case all the triplets defined by the `TripletGeneration`
settings will be scanned for recombination signal.

If the argument is not one of these two arguments, it can be a list of vectors that each contain three sequence names.

For example just two triplets could be scanned:

```{r}
tripletsToScan <- list(c("Seq1", "Seq4", "Seq7"), c("Seq1", "Seq4", "Seq8")) 

MyAnalysis$analyzeSS(tripletsToScan)
```









