---
layout: manualsection
title: Scanning triplets for recombination signal
permalink: 06-manual.html
manual: true
published: false
status: process
---

```{r include=FALSE, echo=FALSE}
FirstGroup <- c("Seq1", "Seq2", "Seq3")
SecondGroup <- c("Seq4", "Seq5", "Seq6")
ThirdGroup <- c("Seq7", "Seq8", "Seq9", "Seq10")
myGroups <- list(FirstGroup, SecondGroup, ThirdGroup) # Put the groups in a list.
MyAnalysis$setParameters("TripletGeneration", Groups = myGroups, PartitionStrictness = 1L)
MyAnalysis
```

This step will use a sliding window to measure pairwise sequence similarity across the informative sites of sequence
triplets. It is started by using the `analyzeSS()` method of the HybRIDS object.
The parameters of the step are:

WindowSize
  : The size in bases of the Sliding Window. The default size is 100.

StepSize
  : The number of bases that the sliding window moves across sequences. The default size is 1. 

The `analyzeSS` method also accepts an argument which tells it which triplets
(of the set you have defined in the `TripletGeneration` step) are to be scanned with the current
`SSAnalysis` settings.

This argument can be set as `"NOT.SCANNED"` (default) in which case, every triplet of the set defined by the 
`TripletGeneration` parameters will be scanned for recombination signal with the current `SSAnalysis`
settings, if it has not previously already been scanned.
This argument can also be set to `"ALL"`, in which case all the triplets defined by the `TripletGeneration`
settings will be scanned for recombination signal.

If the argument is not one of these two arguments, it can be a list of vectors that each contain three sequence names.

For example just two triplets could be scanned:

```{r}
tripletsToScan <- list(c("Seq1", "Seq4", "Seq7"), c("Seq1", "Seq4", "Seq8")) 

MyAnalysis$analyzeSS(tripletsToScan)
```

It is important to note that you can't enter a triplet of names which either contains names which are not names
of your sequences, or is not allowed by the `TripletGeneration` settings. This is demonstrated in the example below.
Recall from the previous section that the sequences "Seq1", "Seq2", and "Seq3" were defined as being in the same group, and because of the settings, only triplets that compared sequences from different groups were set. Therefore the following will not analyze any triplets, as the selected triplet does not exist. 

```{r}
# This triplet is not allowed, because earlier the TripletGeneration
# settings were set so as only triplets which try to find recombination regions between 
tripletsToScan <- list(c("Seq1", "Seq2", "Seq3"))
MyAnalysis$analyzeSS(tripletsToScan)
```








