---
layout: manualsection
title: Detection of recombination blocks in signal
permalink: 07-manual.html
manual: true
published: false
status: process
---

```{r include=FALSE, echo=FALSE}
FirstGroup <- c("Seq1", "Seq2", "Seq3")
SecondGroup <- c("Seq4", "Seq5", "Seq6")
ThirdGroup <- c("Seq7", "Seq8", "Seq9", "Seq10")
myGroups <- list(FirstGroup, SecondGroup, ThirdGroup) # Put the groups in a list.
MyAnalysis$setParameters("TripletGeneration", Groups = myGroups, PartitionStrictness = 1L)
tripletsToScan <- list(c("Seq1", "Seq4", "Seq7"), c("Seq1", "Seq4", "Seq8"))
MyAnalysis$analyzeSS(tripletsToScan)
```

Having scanned for recombination signal for the desired sequence triplets, the `BlockDetection` step is executed using the `findBlocks` method.

This step identifies regions of recombination in a given sequence triplet by examining the data from the sliding window scans in the previous `SSAnalysis` step. If there are any regions in which two sequences meet a certain sequence similarity threshold, they are treated as potentially recombinant.

The parameters for this step are:

ManualThresholds
  : A vector of numbers between 0 and 100.
  : Each is a percenage sequence similarity threshold.
  : These thresholds are set by the user, and can be used to find recombination,
  : or used as a fallback if HybRIDS fails to automatically decide on some thresholds.
  
AutoThresholds
  : A logical argument (`TRUE`, or `FALSE`), if `TRUE` (default) then during execution of the `findBlocks` method,
  : thresholds will be decided on automatically. If `FALSE` then the `ManualThresholds` will be used.
  
ManualFallback
  : A logical argument (`TRUE`, or `FALSE`), if `TRUE` (default), then if automatic thresholds cannot be found, the
  : `ManualThresholds` will be used instead.
  
SDstringency
  : A numeric value, the higher it is, the closer automatically detected thresholds are allowed to be to the mean
  : sequence similarity across the sequences. It is usually fine to leave this as 2.
  
Like the `analyzeSS` method in the previous section, the `findBlocks` must be given an argument that dictates which triplets of the set defined by the `TripletGeneration` parameters will have blocks found from their scan data.
By default this is set to `"NOT.SEARCHED"` which will find blocks for triplets which have not already had blocks identified from their scan data. This argument can also be set to `"ALL"` or a list of vectors that each contain three sequence names.

The below example finds blocks in the two triplets that were scanned for recombination signal in the previous section:
```{r}
tripletsToSearch <- list(c("Seq1", "Seq4", "Seq7"), c("Seq1", "Seq4", "Seq8"))
MyAnalysis$findBlocks(tripletsToSearch)
```

Note that, as with the `analyzeSS` method, you cannot specify a triplet which was not specified by the `TripletGeneration` settings:

```{r}
tripletsToSearch <- list(c("Seq1", "Seq2", "Seq3"))
MyAnalysis$findBlocks(tripletsToSearch)
```

You also cannot search for blocks for triplets which have not been analyzed by `analyzeSS`:
```{r}
tripletsToSearch <- list(c("Seq1", "Seq4", "Seq9"))
MyAnalysis$findBlocks(tripletsToSearch)
```



